Associations between Chlorophyll \emph{a} and various Microcystin-LR Health Advisory Concentrations

### 2015-12-04 manuscript/manuscript.Rmd
started edits on #6

##Data
We used the 2007 NLA chlorophyll *a* and microcystin-LR concentration data [@usepa2009national]. These data represent a snapshot of water quality from the summer of 2007 for the conterminous United States and were collected as part of an ongoing probabilistic monitoring program [@usepa2009national]. Data on chlorophyll *a* and microcystin-LR concentrations are available for `r nrow(na.omit(nla_dat$CHLA,nla_dat$MCYST_TL_UGL))` lakes.

@@

Conditional probability analysis provides information about the probability of observing one event given another event has also occured.  For this analysis, we used CPA to examine how the conditional probability of exceeding one of the health advisories changes as chlorophyll *a* increases in a lake.  We expect to find higher chlorohpyll *a* concentrations to be associated with higher probabilities of exceeding the microcystin-LR health advisory levels.  We also caclulated bootstrapped 95% confidence intervals (CI) using 1000 bootstrapped samples.  Thus, to identify chlorophyll *a* concentrations of concern we identfiy the value of the upper 95% CI across a range of conditional probabilities of exceeding each health advisory level.  As both microcystin-LR and chlorophyll *a* values were highely skewed right, a log base 10 transformation was used.  Additional details of the specific implementation are available at https://github.com/USEPA/microcystinchla.  A more detailed discussion of CPA is beyond the scope of this paper, but see Paul et al. [-@paul2005development] and Hollister et al. [-@hollister2008cprob] for greater detail.  Lastly, all analyses were conducted using R version 3.2.2 and code and data from this analysis are freely available as an R package at [https://github.com/USAPE/microcystinchla](https://github.com/USAPE/microcystinchla).

#Results

In the 2007 NLA, microcystin-LR concentrations ranged from `r min(nla_dat$MCYST_TL_UGL,na.rm=T)` to `r max(nla_dat$MCYST_TL_UGL, na.rm=T)` $\mu$g/L.  Microcystin-LR concentrations of 0.05 $\mu$g/L represent the detection limits.  Any value greater than that indicates the presence of microcystin-LR.  Of those lakes with microcystin, the median concentration was `r round(median(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T), 2)` and the mean was `r round(mean(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T), 2)`.  Of all lakes sampled, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=0.3])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the USEPA Child level, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1.6])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the USEPA Adult level,`r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO Drinking level,and `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=2])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO Recreational level. For chlorophyll *a*, the range was `r min(nla_dat$CHLA,na.rm=T)` to `r max(nla_dat$CHLA, na.rm=T)` $\mu$g/L.  All lakes had reported chlorophyll *a* concentrations that exceeded detection limits.  The median concentration was `r median(nla_dat$CHLA,na.rm=T)` $\mu$g/L and the mean was `r round(mean(nla_dat$CHLA,na.rm=T),2)` $\mu$g/L.  The associations between chlorophyll *a* and the upper confidence interval across a range of conditional probability values is shown in Table \ref{tab:mc_chla_table}.  Specific chlorophyll *a* that are associated with greater than even odds of exceeding the advisory levels were `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, and `r mc_chla[4]` for 0.3, 1.0, 1.6 and 2.0 $\mu$g/L advisory levels, respectively (Table \ref{tab:mc_chla_table} & Figure \ref{fig:multi_cp_plot}).

@@

When using the chlorophyll *a* cutoffs as an indicator of microcystin-LR exceedances, the error that should be avoided is predicting that no exceedance has occurred when in fact it has.  In other words, we would like to avoid Type II errors and minimize the proportion of false negatives.  For the four chlorophyll *a* cut-offs we had a proportion of false negatives of `r fn_percent$epa_child*100`%, `r fn_percent$who_drink*100`%, `r fn_percent$epa_adult*100`% and , `r fn_percent$who_rec*100`% for the U.S EPA childrens drinking water, the WHO drinking water, the USEPA adult drinking water, and the WHO recreational advisories, respectively.  In each case we miss less than 10% of the lakes that are in fact exceeding the microcystin-LR advisory.

@@

#Acknowledgements

>>>>

##Data
We used the 2007 NLA chlorophyll *a* and microcystin-LR concentration data [@usepa2009national, [NLA](http://www2.epa.gov/national-aquatic-resource-surveys/national-lakes-assessment)]. These data represent a snapshot of water quality from the summer of 2007 for the conterminous United States and were collected as part of an ongoing probabilistic monitoring program [@usepa2009national]. Data on chlorophyll *a* and microcystin-LR concentrations are available for `r nrow(na.omit(nla_dat$CHLA,nla_dat$MCYST_TL_UGL))` lakes.

@@

Conditional probability analysis provides information about the probability of observing one event given another event has also occured.  For this analysis, we used CPA to examine how the conditional probability of exceeding one of the health advisories changes as chlorophyll *a* increases in a lake.  We expect to find higher chlorohpyll *a* concentrations to be associated with higher probabilities of exceeding the microcystin-LR health advisory levels.  We also caclulated bootstrapped 95% confidence intervals (CI) using 1000 bootstrapped samples.  Thus, to identify chlorophyll *a* concentrations of concern we identfiy the value of the upper 95% CI across a range of conditional probabilities of exceeding each health advisory level.  Using the upper confidence limit to identify a threshold is justified as it ensures that a given threshold is unlikely to miss a microcystin exceedance.

As both microcystin-LR and chlorophyll *a* values were highely skewed right, a log base 10 transformation was used.  Additional details of the specific implementation are available at https://github.com/USEPA/microcystinchla.  A more detailed discussion of CPA is beyond the scope of this paper, but see Paul et al. [-@paul2005development] and Hollister et al. [-@hollister2008cprob] for greater detail.  Lastly, all analyses were conducted using R version 3.2.2 and code and data from this analysis are freely available as an R package at [https://github.com/USAPE/microcystinchla](https://github.com/USAPE/microcystinchla).

#Results
In the 2007 NLA, microcystin-LR concentrations ranged from `r min(nla_dat$MCYST_TL_UGL,na.rm=T)` to `r max(nla_dat$MCYST_TL_UGL, na.rm=T)` $\mu$g/L.  Microcystin-LR concentrations of 0.05 $\mu$g/L represent the detection limits.  Any value greater than that indicates the presence of microcystin-LR.  Of those lakes with microcystin, the median concentration was `r round(median(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T), 2)` and the mean was `r round(mean(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T), 2)`.  Of all lakes sampled, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=0.3])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the USEPA Child level, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1.6])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the USEPA Adult level,`r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO Drinking level,and `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=2])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO Recreational level. For chlorophyll *a*, the range was `r min(nla_dat$CHLA,na.rm=T)` to `r max(nla_dat$CHLA, na.rm=T)` $\mu$g/L.  All lakes had reported chlorophyll *a* concentrations that exceeded detection limits.  The median concentration was `r median(nla_dat$CHLA,na.rm=T)` $\mu$g/L and the mean was `r round(mean(nla_dat$CHLA,na.rm=T),2)` $\mu$g/L.  The associations between chlorophyll *a* and the upper confidence interval across a range of conditional probability values is shown in Table \ref{tab:mc_chla_table}.  Specific chlorophyll *a* that are associated with greater than even odds of exceeding the advisory levels were `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, and `r mc_chla[4]` for 0.3, 1.0, 1.6 and 2.0 $\mu$g/L advisory levels, respectively (Table \ref{tab:mc_chla_table} & Figure \ref{fig:multi_cp_plot}).

@@

When using the chlorophyll *a* cutoffs as an indicator of microcystin-LR exceedances, the error that should be avoided is predicting that no exceedance has occurred when in fact it has.  In other words, we would like to avoid Type II errors and minimize the proportion of false negatives.  For the four chlorophyll *a* cut-offs we had a proportion of false negatives of `r fn_percent$epa_child*100`%, `r fn_percent$who_drink*100`%, `r fn_percent$epa_adult*100`% and , `r fn_percent$who_rec*100`% for the U.S EPA childrens drinking water, the WHO drinking water, the USEPA adult drinking water, and the WHO recreational advisories, respectively.  In each case we miss less than 10% of the lakes that are in fact exceeding the microcystin-LR advisory.  While this method performs well with regard to the false negative percentag, it is possible that is a relic of the NLA dataset and testing with additional data should be explore to further substatiate this result.

@@

#Acknowledgements
We would like to thank Anne Kuhn, Bryan Milstead, John Kiddon, Joe LiVolsi, Tim Gleason, and Wayne Munns for constructive reviews of this paper. This paper has not been subjected to Agency review. Therefore, it does not necessary reflect the views of the Agency. Mention of trade names or commercial products does not constitute endorsement or recommendation for use. This contribution is identified by the tracking number ORD-XXXXX of the Atlantic Ecology Division, Office of Research and Development, National Health and Environmental Effects Research Laboratory, US Environmental Protection Agency.

### 2015-12-04 manuscript/manuscript.tex
started edits on #6

We used the 2007 NLA chlorophyll \emph{a} and microcystin-LR
concentration data (USEPA 2009). These data represent a snapshot of
water quality from the summer of 2007 for the conterminous United States
and were collected as part of an ongoing probabilistic monitoring
program (USEPA 2009). Data on chlorophyll \emph{a} and microcystin-LR
concentrations are available for lakes.

@@

Conditional probability analysis provides information about the
probability of observing one event given another event has also occured.
For this analysis, we used CPA to examine how the conditional
probability of exceeding one of the health advisories changes as
chlorophyll \emph{a} increases in a lake. We expect to find higher
chlorohpyll \emph{a} concentrations to be associated with higher
probabilities of exceeding the microcystin-LR health advisory levels. We
also caclulated bootstrapped 95\% confidence intervals (CI) using 1000
bootstrapped samples. Thus, to identify chlorophyll \emph{a}
concentrations of concern we identfiy the value of the upper 95\% CI
across a range of conditional probabilities of exceeding each health
advisory level. As both microcystin-LR and chlorophyll \emph{a} values
were highely skewed right, a log base 10 transformation was used.
Additional details of the specific implementation are available at
\url{https://github.com/USEPA/microcystinchla}. A more detailed
discussion of CPA is beyond the scope of this paper, but see Paul et al.
(2005) and Hollister et al. (2008) for greater detail. Lastly, all
analyses were conducted using R version 3.2.2 and code and data from
this analysis are freely available as an R package at
\href{https://github.com/USAPE/microcystinchla}{\url{https://github.com/USAPE/microcystinchla}}.

@@

When using the chlorophyll \emph{a} cutoffs as an indicator of
microcystin-LR exceedances, the error that should be avoided is
predicting that no exceedance has occurred when in fact it has. In other
words, we would like to avoid Type II errors and minimize the proportion
of false negatives. For the four chlorophyll \emph{a} cut-offs we had a
proportion of false negatives of 9\%, 8\%, 6\% and , 5\% for the U.S EPA
childrens drinking water, the WHO drinking water, the USEPA adult
drinking water, and the WHO recreational advisories, respectively. In
each case we miss less than 10\% of the lakes that are in fact exceeding
the microcystin-LR advisory.

@@

>>>>

We used the 2007 NLA chlorophyll \emph{a} and microcystin-LR
concentration data (USEPA 2009,
\href{http://www2.epa.gov/national-aquatic-resource-surveys/national-lakes-assessment}{NLA}).
These data represent a snapshot of water quality from the summer of 2007
for the conterminous United States and were collected as part of an
ongoing probabilistic monitoring program (USEPA 2009). Data on
chlorophyll \emph{a} and microcystin-LR concentrations are available for
lakes.

@@

Conditional probability analysis provides information about the
probability of observing one event given another event has also occured.
For this analysis, we used CPA to examine how the conditional
probability of exceeding one of the health advisories changes as
chlorophyll \emph{a} increases in a lake. We expect to find higher
chlorohpyll \emph{a} concentrations to be associated with higher
probabilities of exceeding the microcystin-LR health advisory levels. We
also caclulated bootstrapped 95\% confidence intervals (CI) using 1000
bootstrapped samples. Thus, to identify chlorophyll \emph{a}
concentrations of concern we identfiy the value of the upper 95\% CI
across a range of conditional probabilities of exceeding each health
advisory level. Using the upper confidence limit to identify a threshold
is justified as it ensures that a given threshold is unlikely to miss a
microcystin exceedance.

As both microcystin-LR and chlorophyll \emph{a} values were highely
skewed right, a log base 10 transformation was used. Additional details
of the specific implementation are available at
\url{https://github.com/USEPA/microcystinchla}. A more detailed
discussion of CPA is beyond the scope of this paper, but see Paul et al.
(2005) and Hollister et al. (2008) for greater detail. Lastly, all
analyses were conducted using R version 3.2.2 and code and data from
this analysis are freely available as an R package at
\href{https://github.com/USAPE/microcystinchla}{\url{https://github.com/USAPE/microcystinchla}}.

@@

When using the chlorophyll \emph{a} cutoffs as an indicator of
microcystin-LR exceedances, the error that should be avoided is
predicting that no exceedance has occurred when in fact it has. In other
words, we would like to avoid Type II errors and minimize the proportion
of false negatives. For the four chlorophyll \emph{a} cut-offs we had a
proportion of false negatives of 9\%, 8\%, 6\% and , 5\% for the U.S EPA
childrens drinking water, the WHO drinking water, the USEPA adult
drinking water, and the WHO recreational advisories, respectively. In
each case we miss less than 10\% of the lakes that are in fact exceeding
the microcystin-LR advisory. While this method performs well with regard
to the false negative percentag, it is possible that is a relic of the
NLA dataset and testing with additional data should be explore to
further substatiate this result.

@@

We would like to thank Anne Kuhn, Bryan Milstead, John Kiddon, Joe
LiVolsi, Tim Gleason, and Wayne Munns for constructive reviews of this
paper. This paper has not been subjected to Agency review. Therefore, it
does not necessary reflect the views of the Agency. Mention of trade
names or commercial products does not constitute endorsement or
recommendation for use. This contribution is identified by the tracking
number ORD-XXXXX of the Atlantic Ecology Division, Office of Research
and Development, National Health and Environmental Effects Research
Laboratory, US Environmental Protection Agency.
